Causal inference and co-evolutionary contingencies using dynamic phylogenetic models

### 2022-07-07 manuscript.qmd
simulations of phylogenetic confounding

@@

To understand when the evolutionary regression coefficient can be interpreted as a causal effect, we must first clarify what "causality" means. Following Pearl (some citations), we say that $X$ is a cause of $Y$ when $Pr(Y|do(X = x)) \neq Pr(Y|do(X = \textrm{not x}))$. The do() operator represents an *intervention* in which the value of some variable is fixed (for example, through assignment in a randomized experiment). If the probability distribution of $Y$ changes based on the value of $X$ that is assigned, then $X$ has a causal effect on $Y$. Note here that $X$ refers to some variable while $x$ refers to some unit value of that variable.

@@

Phylogenetic generalized multilevel (or mixed-effects) models (PGLMMs) extend the study of trait coevolution by leveraging analytic strategies developed in quantitative genetics to accommodate more complex, multivariate model designs with non-Gaussian outcomes [@hadfield2010general; @ives2010phylogenetic; @ives2011generalized] <!--# there's also Lynch 1991 METHODS FOR THE ANALYSIS OF COMPARATIVE DATA IN EVOLUTIONARY BIOLOGY, Evolution -->. Rather than explicitly scaling the off-diagonals of $\boldsymbol{V}_s$ with some parameter $\lambda$, phylogenetic signal is quantified as the ratio of the variance of phylogenetic random effects $\nu$ (also called 'mixed' or 'varying' effects) to total trait variance. For example, an observation of a Gaussian trait $Y$ from a species or population $j$ and predictor $X$ is modeled as:

@@

>>>>

TWO different reasons to "adjust for phylogeny"

(1) Repeated measures of closely-related species, non-independence (std errors too small)

(2) Clustering in multivariate data can lead to confounding due to sampling variation. Type I (between-cluster sampling variation leads to spurious correlation) and Type II error (between cluster sampling variation leads to underestimated correlation).

Why would we care about the phylogenetic correlation? Because this is an indicator that the traits co-evolve across the phylogenetic tree rather than just at the tips.

@@

To understand when the evolutionary regression coefficient can be interpreted as a causal effect, we must first clarify what "causality" means. Following Pearl (some citations), we say that $X$ is a cause of $Y$ when $Pr(Y|do(X = x)) \neq Pr(Y|do(X = \textrm{not x}))$. The do() operator represents an *intervention* in which the value of some variable is fixed (for example, through assignment in a randomized experiment). If the probability distribution of $Y$ is sensitive to the value of $X$ that is assigned, then $X$ has a causal effect on $Y$. Note here that $X$ refers to a random variable while $x$ refers to some realized unit value of that variable.

@@

Phylogenetic generalized multilevel (or mixed-effects) models (PGLMMs) extend the study of trait coevolution by leveraging analytic strategies developed in quantitative genetics to accommodate more complex, multivariate model designs with non-Gaussian outcomes [@lynch1991methods; @hadfield2010general; @ives2010phylogenetic; @ives2011generalized] <!--# there's also Lynch 1991 METHODS FOR THE ANALYSIS OF COMPARATIVE DATA IN EVOLUTIONARY BIOLOGY, Evolution -->. Rather than explicitly scaling the off-diagonals of $\boldsymbol{V}_s$ with some parameter $\lambda$, phylogenetic signal is quantified as the ratio of the variance of phylogenetic random effects $\nu$ (also called 'mixed' or 'varying' effects) to total trait variance. For example, an observation of a Gaussian trait $Y$ from a species or population $j$ and predictor $X$ is modeled as:

@@

#### Multivariate PGLMMs

Residual and phylogenetic correlations can be different! Why? Plasticity and/or microevolutionary response to X may not be the same as the macroevolutionary process, a violation of *ergodicity*. or correlated measurement error.
